---
title: "Notes 5 - Databases and SQL"
author: |
  | Rick Brown
  | Southern Utah University
date: "Math 3190"
output: 
 beamer_presentation:
    theme: CambridgeUS
header-includes:
  - \usepackage{amsmath}
  - \makeatletter
  - \preto{\@verbatim}{\topsep=-10pt \partopsep=4pt }
  - \makeatother
tables: true
urlcolor: blue
editor_options: 
  chunk_output_type: console
---

\tableofcontents

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DBI)
library(RMySQL)
library(tidyverse)
con = dbConnect(RMySQL::MySQL(),
               dbname='sakila',
               host='localhost',
               user='root',
               password='mysql_root')
con <- DBI::dbConnect(RSQLite::SQLite(),
                      dbname=paste0("~/My Drive/Math 3190 - Fundamentals ",
                                    "of Data Science/Spring 2024/GitHub",
                                    "/Math3190_Sp24/Data/sakila.sqlite")
                      )
```

# Databases and SQL

## Intro to Database Design

In many businesses and government agencies, data files are not directly accessible. Instead, data are stored in databases that are usually in the cloud. 
The most common type of database is a *relational database*. These types of databases are made up of **tables**, which are very similar to data frames in **R**. These tables usually have at least one variable in common with at least one other table that can be used to connect them. 

## Relational Database Example

\begin{center}
\includegraphics[width=4.5in]{sql_figs/sakila.png}
\end{center}

## SQL Intro

The most common way to access information stored in relational databases is with SQL.

SQL stands for \underline{structured query language} and is ubiquitous in business, government, and academia where large data sets are used. SQL is usually either pronounced like "sequel" or the letters are individually said "S-Q-L". 

We are going to go over the common SQL commands and then look at how to access SQL databases and use SQL within **R**.

SQL is one of the most basic "coding languages" there is because of how similar the statements are to spoken English.

## `SELECT`, `FROM`, and `WHERE` Statements

The backbone of SQL are the `SELECT`, `FROM`, and `WHERE` statements.

* `SELECT` is very much like the `select()` function in the tidyverse (`dplyr` package). It selects the columns or variables from the database. You can use "*" to select all columns.

* `FROM` is the statement that specifies the table in the database from which we are taking the data.

* `WHERE` is similar to the `filter()` function in the tidyverse. This will subset the table based on some criterion or criteria. 

Note: these words do not *need* to be written in all capital letters, but it is the convention to type them in all caps for readability. 

## SQL Syntax

The basic syntax for SQL statements is as follows:

```{sql eval = F, echo = T}
SELECT column1, column2, ...
FROM table_name
WHERE condition;
```




## SQL Example

Suppose we have a database named `sakila` that models a DVD rental store, featuring things like films, actors, film-actor relationships. In this database are many tables, one of which is "film". Here is some information about the film table:

\vspace{-.2in}
\phantom{.}\hspace{2.5in}\includegraphics[width=1.9in]{sql_figs/sakila_film.pdf}
\vspace{-2.1in}

Let's run some SQL queries with\
this table.

\vspace{2in}
\phantom{.}

## SQL Query 1

```{sql connection=con, echo = T, max.print = 5, tab.cap = "My Caption", output.var = "pg_movies", eval = T}
SELECT title, release_year, rating
FROM film
WHERE rating = 'PG';
```
```{sql connection=con, echo = F, max.print = 5, tab.cap = "My Caption"}
SELECT title, release_year, rating 
FROM film
WHERE rating = 'PG';
```

## WHERE Statement Commands

The following can be used in a `WHERE` statement:

\begin{table}
\centering
\begin{tabular}{|c|c|c|}
\hline
Operator & Description\\
\hline
$=$ & Equal\\
$>$ & Greater than\\
$<$ & Less than\\
$>=$ & Greater than or equal\\
$<=$ & Less than or equal\\
$<>$ or $!=$ & Not equal\\
AND & Restricts the output based on another criterion.\\
OR & Expands the output based on another criterion.\\
NOT & Gives the opposite of the criterion.\\
BETWEEN & Between a certain range.\\
IN & To specify multiple possible values for a column.\\
LIKE & Search for a pattern.\\
\hline
\end{tabular}
\end{table}

## SQL Query 2: BETWEEN

The syntax for the `BETWEEN` statement is
```{sql connection=con, echo = T, eval = F, max.print = 5}
WHERE column_name BETWEEN value1 AND value2
```

```{sql connection=con, echo = T, max.print = 5}
SELECT film_id, title, length, rental_duration
FROM film
WHERE film_id BETWEEN 13 AND 14 
  OR film_id BETWEEN 21 AND 23;
```

## SQL Query 3: IN
The `IN` operator is a shorthand for multiple `OR` conditions.

```{sql connection=con, echo = T, max.print = 4}
SELECT film_id, title, length, rental_duration
FROM film
WHERE film_id IN ("1", "50", "100", "613");
```

Note: the quotes aren't needed for numbers, but can still be used.

## SQL Query 4: LIKE
There are two wildcards often used in conjunction with the `LIKE` operator:

* The percent sign `%` represents zero, one, or multiple characters.
* The underscore sign `_` represents one single character. This can be repeated to stand for multiple characters. 

```{sql connection=con, echo = T, max.print = 4}
SELECT film_id, title, length, rental_duration
FROM film
WHERE title LIKE "b%" AND length >= 175;
```

## SQL Query 5: LIKE

```{sql connection=con, echo = T, max.print = 5}
SELECT film_id, title, length, rental_duration
FROM film
WHERE title LIKE "__ab%";
```

## More on SELECT Statments

In the `SELECT` statement, instead of simply putting column names, we can also use the following functions:

* `MIN()` returns the minimum value of a numeric column.
* `MAX()` returns the maximum value of a numeric column.
* `COUNT()` returns the number of rows that matches a specified criterion.
* `SUM()` returns the total sum of a numeric column.
* `AVG()` returns the average value of a numeric column.

Additionally, we can use aliases to rename columns with the `AS` statement.

## SQL Query 6: SELECT Functions

```{sql connection=con, echo = T, max.print = 5}
SELECT AVG(length) AS "Average Length"
FROM film;
```


## SQL Query 7: SELECT Functions

```{sql connection=con, echo = T, max.print = 5}
SELECT COUNT(title) AS "Number of A or B Movies"
FROM film
WHERE title LIKE "a%" OR title LIKE "b%";
```

## SQL Query 8: SELECT Functions

The following will return all the films with a length higher than the average length.

```{sql connection=con, echo = T, max.print = 5}
SELECT title, length
FROM film
WHERE length > (SELECT AVG(length) FROM film);
```

## ORDER BY Statement 

We can also use the `ORDER BY` statement to change the order in which the data table is given.

The syntax for `ORDER BY` is 
```{sql connection=con, echo = T, eval = F, max.print = 5}
ORDER BY column1, column2, .. columnN
```

We can put a a `ASC` for ascending or `DESC` for descending values after any one of the columns we are ordering by. 

## SQL Query 9: ORDER BY Statement

```{sql connection=con, echo = T, max.print = 5}
SELECT title, length
FROM film
WHERE length > (SELECT AVG(length) FROM film)
ORDER BY length ASC, title DESC;
```


## GROUP BY Statement 

The `GROUP BY` statement groups rows that have the same values into summary rows. It is often used with aggregate functions (`COUNT()`, `MAX()`, `MIN()`, `SUM()`, `AVG()`) to group the result-set by one or more columns.

Syntax:
```{sql eval = F, echo = T}
SELECT column_name(s)
FROM table_name
WHERE condition
GROUP BY column_name(s)
ORDER BY column_name(s);
```

## SQL Query 10: GROUP BY Statement

```{sql connection=con, echo = T, max.print = 5}
SELECT AVG(length), rating
FROM film
WHERE release_year = 2006
GROUP BY rating
ORDER BY rating;
```

## SELECT DISTINCT Statement

The `SELECT DISTINCT` statement is used to return only distinct (different) values.

Syntax:
```{sql connection=con, eval = F, echo = T, max.print = 5}
SELECT DISTINCT column1, column2, ...
FROM table_name;
```

## SQL Query 11: DISTINCT Statement

```{sql connection=con, echo = T, max.print = 5}
SELECT DISTINCT release_year
FROM film;
```

So this database only has films from the year 2006.


## UNION Statement

The `UNION` operator is used to combine the result-set of two or more `SELECT` statements.

Syntax:
```{sql eval = F, echo = T}
SELECT column_name(s) FROM table1
UNION
SELECT column_name(s) FROM table2;
```

## SQL Query 12: UNION Statement

```{sql connection=con, echo = T, max.print = 5}
SELECT AVG(length) AS "Average Length and Payment Amount"
FROM film
UNION 
SELECT AVG(amount)
FROM payment;
```

## JOIN Statements

A `JOIN` clause is used to combine rows from two or more tables, based on a related column between them. There are four types of `JOIN` statements:

* `(INNER) JOIN`: Returns records that have matching values in both tables.
* `LEFT (OUTER) JOIN`: Returns all records from the left table, and the matched records from the right table. 

\begin{center}
\includegraphics[width = 1.5in]{sql_figs/inner_join.png}
\includegraphics[width = 1.5in]{sql_figs/left_join.png}
\end{center}

## JOIN Statements

* `RIGHT (OUTER) JOIN`: Returns all records from the right table, and the matched records from the left table.
* `FULL (OUTER) JOIN`: Returns all records when there is a match in either left or right table.

\begin{center}
\includegraphics[width = 1.5in]{sql_figs/right_join.png}
\includegraphics[width = 1.5in]{sql_figs/full_outer_join.png}
\end{center}

## JOIN Statements

The syntax for JOIN statements is as follows:

```{sql, echo = T, eval = F}
SELECT column_name(s)
FROM table1
(JOIN STATEMENT) table2
  ON table1.column_name = table2.column_name;
```

Replace the `(JOIN STATEMENT)` with `INNER JOIN`, `LEFT JOIN`, `RIGHT JOIN`, or `FULL JOIN`.

You can select columns from either table, so it is common to denote which table we are selecting from by using `table.column_name`.

## JOIN Examples

Consider the following two tables called Student and StudentCourse, respectively. 

\includegraphics[width = 2.86in]{sql_figs/join_ex_tab1.png}
\includegraphics[width = 1.63in]{sql_figs/join_ex_tab2.png}

## JOIN Example: INNER JOIN 

```{sql connection=con, echo = T, eval = F}
SELECT StudentCourse.COURSE_ID, Student.NAME, Student.AGE 
FROM Student
INNER JOIN StudentCourse
  ON Student.ROLL_NO = StudentCourse.ROLL_NO;
```
\includegraphics[width = 3in]{sql_figs/join_ex_innerjoin.png}

## JOIN Example: LEFT JOIN 

```{sql connection=con, echo = T, eval = F}
SELECT Student.NAME, StudentCourse.COURSE_ID 
FROM Student
LEFT JOIN StudentCourse 
ON StudentCourse.ROLL_NO = Student.ROLL_NO;
```
\includegraphics[width = 2.5in]{sql_figs/join_ex_leftjoin.png}

## JOIN Example: RIGHT JOIN 

```{sql connection=con, echo = T, eval = F}
SELECT Student.NAME,StudentCourse.COURSE_ID 
FROM Student
RIGHT JOIN StudentCourse 
ON StudentCourse.ROLL_NO = Student.ROLL_NO;
```
\includegraphics[width = 2.5in]{sql_figs/join_ex_rightjoin.png}

## JOIN Example: FULL JOIN 

```{sql connection=con, echo = T, eval = F}
SELECT Student.NAME,StudentCourse.COURSE_ID 
FROM Student
FULL JOIN StudentCourse 
ON StudentCourse.ROLL_NO = Student.ROLL_NO;
```
\includegraphics[width = 1.9in]{sql_figs/join_ex_fulljoin.png}

# SQL in R Markdown

## SQL in R Markdown

In R Markdown, you can insert an SQL code chunk. 

````{verbatim, echo = T}
```{sql connection=}
```
````
We will need to establish a connection before using it, though. This connection must be to a database, either remote or local. 

## Define an SQL Connection in **R**

To connect to a remote database, we can use the following code chunk in R Markdown. We will need a few SQL and database packages, though. Some common ones are `DBI`, `RMySQL`, and `RSQL`, and `RSQLite`. This database is a MySQL database, so we will need `RMySQL`. 

````{verbatim, echo = T}
```{r, echo = T, eval = T}
library(DBI)
library(RMySQL)
con <- dbConnect(RMySQL::MySQL(),
                 dbname = "sakila",
                 host = "relational.fit.cvut.cz",
                 port = 3306,
                 user = "guest",
                 password = "relational")
```
````


## Define an SQL Connection in **R** 

Note: this website for the sakilla database has since gone down since these notes were created. A remote database I found that is still up (as of January 31, 2024) can be accessed using the code on the following slide:

````{verbatim, echo = T}
```{r, echo = T, eval = T}
con = dbConnect(RMySQL::MySQL(),
                dbname='Rfam',
                host='mysql-rfam-public.ebi.ac.uk',
                port=4497,
                user='rfamro',
                password='')
```
````
The following slides will still use the sakila database as an example.


## Define an SQL Connection in **R**

Then we can use that connection in our SQL code chunk. 

````{verbatim, echo = T}
```{sql connection = con, echo = T, eval = T, max.print = 5}
SELECT title, release_year, rating
FROM film
WHERE rating = 'PG';
```
````

```{sql connection = con, echo = F, eval = T, max.print = 5}
SELECT title, release_year, rating
FROM film
WHERE rating = 'PG';
```

## Creating a Data Frame in **R** from SQL

Instead of printing the table, we can store the table we created from our SQL command as a data frame using the `output.var` argument in the SQL code chunk:

````{verbatim, echo = T}
```{sql connection = con, output.var = "pg_movies"}
SELECT title, release_year, rating
FROM film
WHERE rating = 'PG';
```
````


## Using SQL Table in **R**
Now we can access the `pg_movies` variable in **R**.
```{r, echo = T}
pg_movies <- tibble(pg_movies)
head(pg_movies, 5)
dim(pg_movies)
```

## Using Tidyverse Functions

The same thing can be accomplished if we had read in the film table as a data frame or tibble. 
```{r, echo = T, eval = F}
movies <- read.csv("film.csv")
pg_movies <- movies |>
  filter(rating == "PG") |>
  select(title, release_year, rating) |> tibble()
head(pg_movies, 5)
```
```{r, echo = F, eval = T}
movies <- read.csv(paste0("~/My Drive/Math 3190 - Fundamentals of Data Science",
"/Data/Sakila DB Data/film.csv"))
pg_movies <- movies |>
  filter(rating == "PG") |>
  select(title, release_year, rating) |> tibble()
head(pg_movies, 5)
```

## Using Tidyverse Functions

The following statements and functions are equivalent:

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
SQL & {\bf R} Tidyverse\\
\hline
SELECT & select()\\
WHERE & filter()\\
ORDER BY & arrange()\\
GROUP BY & group\_by()\\
INNER JOIN & inner\_join()\\
LEFT JOIN & left\_join()\\
RIGHT JOIN & right\_join()\\
FULL (OUTER) JOIN & full\_join()\\
\hline
\end{tabular}
\end{center}


## Using **R** Objects in SQL Chunks

If you need to use the values of an **R** variables into SQL queries, you can do so by prefacing **R** variable references with a ?. For example:
````{verbatim, echo = T}
```{r}
movie_length = 130
```

```{sql, connection=con, max.print = 5}
SELECT title, length
FROM film 
WHERE length >= ?movie_length
```
````

## Using **R** Objects in SQL Chunks

```{r, echo = F, eval = T}
movie_length = 130
```
```{sql, connection=con, echo = F, eval = T, max.print = 5}
SELECT title, length
FROM film 
WHERE length >= ?movie_length
```

## Close Session

When we are finished with connecting to the database, it is best to close the connection using the following line:

```{r, echo = T, eval = T}
dbDisconnect(con) # Disconnects from server.
```


## Other SQL Uses

There are lots of other things we can do using SQL. We can create tables, create databases, drop tables, insert rows and columns into tables, and more.

For this class, however, we will leave it there since most of the time, we will simply be accessing databases for data analysis rather than editing them. 


# Local Databases

## Simulating SQL Database with Local Files in Memory

To simulate a connection to a database, we can use the following:

\footnotesize
```{r establish memory db, eval = T, echo = T}
con <- dbConnect(RSQLite::SQLite(), ":memory:")
sakila_files <- list.files(paste0("~/My Drive/Math 3190 - Fundamentals ",
                                  "of Data Science/Data/Sakila DB Data"))
for (file in sakila_files) {
  data_file <- read.csv(paste0("~/My Drive/Math 3190 - Fundamentals ",
                               "of Data Science/Data/Sakila DB Data/", 
                               file))
  # Add data frame as a table                             
  dbWriteTable(con, substr(file, 1, nchar(file) - 4), data_file) 
  rm(data_file)
}
dbListTables(con)
```
\vspace{.05in}

\normalsize
These tables are not stored in **R**'s memory, though! 

## Simulating SQL Database with Local Files

```{sql connection=con, max.print = 5}
SELECT title, length
FROM film
WHERE length > (SELECT AVG(length) FROM film);
```

## Connect to Local Data Databases

If we have an SQL-type file (like one that ends in `.sqlite`), then we can connect to it without reading it into **R**'s memory. 

For example, I have a database called "portal_mammals.sqlite" (downloadable from [here](https://ndownloader.figshare.com/files/2292171)). If it is in my working directory, I can connect to this database using\
\

<!-- Tutorial link using this database: https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html -->

```{r mammals1}
mammals_con <- DBI::dbConnect(RSQLite::SQLite(), 
                              "portal_mammals.sqlite")
DBI::dbListTables(mammals_con) # List all database tables
```

## Connect to Local Data Databases

Then I can run SQL queries using the `dbGetQuery()` function:
```{r, eval = T, echo = T}
dbGetQuery(mammals_con, "SELECT * FROM surveys;") |> head()
```

or I can connect to the database in R Markdown using
````{verbatim, echo = T}
```{sql connection=mammals_con, max.print = 5}

```
````

## Connect to Local Data Databases

Select from December 2000:
```{sql connection=mammals_con, max.print = 5}
SELECT record_id, month, day, year, species_id, sex, weight
FROM surveys
WHERE month = 12 AND year = 2000;
```


## Create a Local Database

We can create a .sqlite file from a .csv using the `csv_to_sqlite()` function from the `inborutils` package. This is not available on the CRAN, but can be installed using `devtools::install_github("inbo/inborutils")`. 

This function reads a csv in smaller chucks (50,000 lines at a time by default) and writes it to a .sqlite file so we can connect to it as a database. 

Example:
\vspace{-.05in}
```{r, eval = F, echo = T}
library(inborutils)
csv_to_sqlite(csv_file = "combined.csv",
              sqlite_file = "combined.sqlite", 
              table_name = "combined")
              
combined_con <- DBI::dbConnect(RSQLite::SQLite(), 
                               "combined.sqlite")
dbGetQuery(combined_con, "SELECT * FROM combined") |> head()
```


## Create a Local Database

```{r, eval = T, echo = F, message = F}
options(width = 60)
library(inborutils)
csv_to_sqlite(csv_file = "~/My Drive/Math 3190 - Fundamentals of Data Science/Spring 2024/GitHub/data_raw/combined.csv",
              sqlite_file = "~/My Drive/Math 3190 - Fundamentals of Data Science/Spring 2024/GitHub/data_raw/combined.sqlite", 
              table_name = "combined")
              
combined_con <- DBI::dbConnect(RSQLite::SQLite(), "~/My Drive/Math 3190 - Fundamentals of Data Science/Spring 2024/GitHub/data_raw/combined.sqlite")
dbGetQuery(combined_con, "SELECT * FROM combined") |> head()
```


## Create a Local Database

You can add another table to the same .sqlite file using `csv_to_sqlite()` as long as a different table name is specified.

```{r, eval = F, echo = T, message = F}
csv_to_sqlite(csv_file = "actor.csv",
              sqlite_file = "sakila.sqlite", 
              table_name = "actor")
csv_to_sqlite(csv_file = "film.csv",
              sqlite_file = "sakila.sqlite", 
              table_name = "film")
```

The above code will create the `sakila.sqlite` file in the first function call with the table "actor". The second function call will then add the "film" table to that same file.


## Convert from dplyr Functions to SQL

The `dbplyr` package is an add-on to the `dplyr` package that allows you to use remote database tables as if they are in-memory data frames by automatically converting dplyr code into SQL. We can install it in the normal way: `install.packages("dbplyr")`. If you don't install it, you will be prompted to when you do something that requires it. 

Example:

\small
```{r}
# Access the SQL table "surveys" from the mammals_con connection
mammal_surveys <- tbl(mammals_con, "surveys")
mammal_surveys
```


## Convert from dplyr Functions to SQL

```{r}
mammal_summary <- mammal_surveys |>
  filter(month == 7 & year == 1992) |>
  group_by(species_id) |>
  summarize(avg_weight = mean(weight, na.rm = TRUE)) |>
  arrange(desc(avg_weight))
head(mammal_summary)
```


## Convert from dplyr Functions to SQL

The `show_query()` function will show an SQL query that will accomplish the same thing the `dplyr` functions did to create the object. Sometimes these queries look a bit strange, but they work! 

```{r}
show_query(mammal_summary)
```


## Same thing with SQL

```{sql connection=mammals_con}
SELECT `species_id`, AVG(`weight`) AS `avg_weight`
FROM (
  SELECT `surveys`.*
  FROM `surveys`
  WHERE (`month` = 7.0 AND `year` = 1992.0)
) AS `q01`
GROUP BY `species_id`
ORDER BY `avg_weight` DESC
```

## Object Sizes

What is especially amazing about this is even though we are working with an object called `mammal_surveys`, it is actually not read into memory. We can see that here using the `obj_size()` function in the `lobstr` package:

```{r object_size}
lobstr::obj_size(mammal_surveys)
# collect function brings the object into memory
mammal_surveys2 <- collect(mammal_surveys) 
lobstr::obj_size(mammal_surveys2)
```
\vspace{.1in}

So, being an **R** wizard also makes you an SQL wizard (by proxy)! 


## Close Connection

As always, we should not forget to disconnect when we are done.
```{r mammals_close, eval = T}
dbDisconnect(con)
dbDisconnect(combined_con)
dbDisconnect(mammals_con)
```

## Conclusion

* Knowing SQL is a great skill to not only access databases remotely, but also to be able to work with large files locally without reading them into **R**. 

* Even though tools exist to convert **R** code to SQL queries, it is still important to know both to be able to handle working with any database or data set.


## Session info

\tiny

```{r session info}
sessionInfo()
```
